# Reproducible Research: Peer Assessment 1

load packages
```{r LoadPackages, echo=TRUE}
library(data.table)
library(ggplot2)
library(xtable)
library(dplyr)
```

## Loading and preprocessing the data
```{r Load_Data, echo=TRUE}
setwd("C:/Users/Nathan/RepData_PeerAssessment1")
list.files()
activity<-read.csv("activity.csv")
head(activity)
summary(activity)
str(activity)
View(activity)
```

Convert the data to a tbl_df  
It's worthwhile to convert large datasets to tbl_df

```{r Convert_to_tbl_df, echo=TRUE}
activity_df<-tbl_df(activity)
activity_df
```

Group the data by date

```{r GroupbyDate, echo=TRUE}
activity_date<-group_by(activity_df,date)
```

Summarise the grouped data by total number of steps taken per day
```{r StepsperDay, echo=TRUE}
stepsPerDay<-summarise(activity_date, sum=sum(steps, na.rm=FALSE))
head(stepsPerDay,n=5)
```

## What is mean total number of steps taken per day?

Calculate the mean and median of total number of steps from "stepsPerDay"
```{r meanSteps, echo=TRUE}
meanSteps<-mean(stepsPerDay$sum, na.rm=TRUE)
meanSteps
medianSteps<-median(stepsPerDay$sum, na.rm=TRUE)
medianSteps
```

Create a Histogram of the total number of steps per day
```{r Histogram, echo=TRUE}
h<-ggplot(stepsPerDay, aes(x=sum))+
        geom_histogram(alpha=1/2,binwidth=1000,fill="darkblue",color="black")
h+ggtitle("Histogram of Total Number of Steps per Day")                      
```

## What is the average daily activity pattern?

Group the data by interval
```{r GroupbyInterval, echo=TRUE}
intervalGrp<-group_by(activity,interval)
head(intervalGrp)
View(intervalGrp)
```

Claculate the average number of steps taken per time interval
```{r AverageStepsPerTimeInterval,echo=TRUE}
AvrgStepsIntvl<-summarise(intervalGrp,Average=mean(steps,na.rm=TRUE))
View(AvrgStepsIntvl)
```

Make a time series plot
```{r TimeSeriePlot, echo=TRUE}
ts<-ggplot(AvrgStepsIntvl, aes(x=interval, y=Average))+geom_line(alpha=1,binwidth=5)
ts+ggtitle("Time Series Plot of the 5-Minute Interval")+xlab("5-Minute Interval")+
        ylab("Average Number of Steps per 5-Minute Interval")
```

## Imputing missing values

Calculate the total number of missing values
```{r MissingValues, echo=TRUE, results='asis'}
summary(activity$steps)
```

input missing values
```{r InputValues, echo=TRUE}
##use the "vIM" package to input missing values
library(VIM)

##Use k-Nearest Neighbour Imputation (kNN) method to imput missing values
activity2<-kNN(activity)
summary(activity2)
```

Create a new data set with missing data filled in
```{r NewDataset2,echo=TRUE}
New_activity<-select(activity2, steps,date,interval)
summary(New_activity)
```

Calculate the New mean and median of total number of steps 
```{r meanSteps2, echo=TRUE}
New_activity_date<-group_by(New_activity,date)
New_stepsPerDay<-summarise(New_activity_date, sum=sum(steps))
New_meanSteps<-mean(New_stepsPerDay$sum)
New_meanSteps
New_medianSteps<-median(New_stepsPerDay$sum)
New_medianSteps
```

Create a NEW Histogram of the total number of steps per day
```{r Histogram2, echo=TRUE}
h<-ggplot(New_stepsPerDay, aes(x=sum))+
        geom_histogram(alpha=1/2,binwidth=1000,fill="green",color="black")
h+ggtitle("Histogram of Total Number of Steps per Day-Without Missing Values")
```

Do these values differ from the estimates from the first part of Assignment  
-Yes, the data with the missing values over estimates the mean and the median as shown below
```{r Comparison, echo=TRUE, results='asis'}
meanSteps
New_meanSteps
medianSteps
New_medianSteps
```

What is the impact of imputing missing values on the estimates  
-Inputing missing values reduces the tendency of over estimating parameters such as the mean,
median, percentiles, and others

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable with "weekday" and "weekend" levels
```{r FactorVariables, echo=TRUE}
##create new variable(day) for day of the week
New_activity1<-New_activity
New_activity1$day<-weekdays(as.Date(New_activity1$date,"%Y-%m-%d"))

##Now create new variable for weekends/weekdays
library(chron)
New_activity2<-New_activity1
New_activity2$week<-is.weekend(New_activity1$date)
New_activity2$week1<-ifelse(New_activity2$week=="FALSE", "Weekday" ,"Weekend")
summary(New_activity2)
New_activity2$week1<-as.factor(New_activity2$week1)
str(New_activity2)
```

Make a panel plot containing a time series plot

Group the new data by interval
```{r GroupbyInterval2, echo=TRUE}
New_intervalGrp<-group_by(New_activity2,interval,week1)
head(New_intervalGrp)
```

Claculate the average number of steps taken per time interval
```{r AverageStepsPerTimeInterval2,echo=TRUE}
New_AvrgStepsIntvl<-summarise(New_intervalGrp,Average=mean(steps))
head(New_AvrgStepsIntvl)
```

Make a time series plot
```{r TimeSeriePlot2, echo=TRUE}
ts<-ggplot(New_AvrgStepsIntvl, aes(x=interval, y=Average))+geom_line(color="blue")
ts+facet_wrap(~week1,nrow=2)+theme(legend.position="none")+
        ggtitle("Time Series Plots by Weekday and Weekend")+ylab("Number of Steps")
```

This is the end of the Assignment.
